Split-Apply-Combine with Dynamic Grouping

### 2023-01-02 paper/article.Rnw
writing

\begin{center}
\begin{tikzpicture}
\tikzstyle{level 1}=[level distance=10mm,sibling distance=40mm]
\tikzstyle{level 2}=[level distance=10mm,sibling distance=10mm]
%  every node/.style = {shape=rectangle, rounded corners,
%    draw, align=center,
%    top color=white, bottom color=blue!20}]]
  \node {\code{01}}
    child { node {\code{011}}
      child {node {\code{0111}}}
      child {node {\code{0112}}} 
      child {node {\code{0113}}} }
    child { node {\code{012}}
      child {node {\code{0121}}}
      child {node {\code{0122}}} 
      child {node {\code{0123}}} 
      child {node {\code{0124}}} };
\end{tikzpicture}
\end{center}

>>>>

\begin{center}
\begin{tikzpicture}
\tikzstyle{level 1}=[level distance=10mm,sibling distance=40mm]
\tikzstyle{level 2}=[level distance=10mm,sibling distance=10mm]
  \node {\code{01}}
    child { node {\code{011}}
      child {node {\code{0111}}}
      child {node {\code{0112}}} 
      child {node {\code{0113}}} }
    child { node {\code{012}}
      child {node {\code{0121}}}
      child {node {\code{0122}}} 
      child {node {\code{0123}}} 
      child {node {\code{0124}}} };
\end{tikzpicture}
\end{center}
Here, the hierarchy suggest to collapsing $\{0111,0112,0113\}$ into $\{011\}$
(when needed) and similar for the right branch. The second level of collapsing
would combine $\{012\}$ with $\{011\}$ into $\{01\}$. The \pkg{accumulate}
package comes with a helper function that creates the collasping scheme from the
lowest-level digits.
<<>>=
nace <- c("0111","0112","0113","0121","0121","0122","0123","0124")
csh_from_digits(nace, levels=2)
@ 
Here, the parameter \code{levels} determines how many collapsing steps will be
computed. Since all codes are prepended with zero, there is no need to collapse
01 any further. The output can be used as argument to the \code{collapse}
parameter of the \code{accumulate()} or \code{cumulate} functions.

The situation becomes a little more involved when hierarchical classifcations
form a tree such that the distance from leave to trunk is not the same for all
leaves (unbalanced tree). This occurs in practice, for example when local
organisations create an extra level of detail for some, but not all leaves.
Below is an example of such a situation.
%
\begin{center}
\begin{tikzpicture}
\tikzstyle{level 1}=[level distance=10mm,sibling distance=50mm]
\tikzstyle{level 2}=[level distance=10mm,sibling distance=10mm]
\tikzstyle{level 2}=[level distance=10mm,sibling distance=15mm]
  \node {\code{01}}
    child { node {\code{011}}
      child {node {\code{0111}}}
      child {node {\code{0112}}} 
      child {node {\code{0113}}} }
    child { node {\code{012}}
      child {node {\code{0121}}}
      child {node {\code{0122}}} 
      child {node {\code{0123}}} 
      child {node {\code{0124}}
       child {node {\code{01241}}}
       child {node {\code{01242}}} } };
\end{tikzpicture}
\end{center}
%
In this case, not all leaves can be collapsed with the same number of
collapsing levels. This provides an issue for specifying the collapsing
sequence as it now depends on the leaf where you start how many collapsing
levels are possible. It also complicates interpretability of the result as the
collapsing level reported in the output, now means different things for
different target groups. The solution chosen in \pkg{accumulate} is to extend
the tree by making copies of the leaves that are not on the lowest level as
follows.
%
\begin{center}
\begin{tikzpicture}
\tikzstyle{level 1}=[level distance=10mm,sibling distance=70mm]
\tikzstyle{level 2}=[level distance=10mm,sibling distance=20mm]
\tikzstyle{level 3}=[level distance=10mm,sibling distance=15mm]
  \node {\code{01}}
    child { node {\code{011}}
      child {node {\code{0111}}
       child {node{\code{0111}}}} 
      child {node {\code{0112}} 
       child {node{\code{0112}}}} 
      child {node {\code{0113}}
       child {node{\code{0113}}}} } 
    child { node {\code{012}}
      child {node {\code{0121}}
       child {node {\code{0121}}}}
      child {node {\code{0122}}
       child {node{\code{0122}}}} 
      child {node {\code{0123}}
       child {node{\code{0123}}}} 
      child {node {\code{0124}}
       child {node {\code{01241}}}
       child {node {\code{01242}}} } };
\end{tikzpicture}
\end{center}
%
The tradeoff is that although there may be some extra calculations in the case
where a leaf is collapsed to itself. The gain is that the specification of the
calculation as well as the interpretation of the results are now uniform
accross all hierarchical classifications. Again, using \code{csh_from_digits()}
deriving the collapsing scheme can be automated.
<<>>=
nace <- c("0111","0112","0113","0121","0122","0123","01241","01242")
csh_from_digits(nace,levels=3)
@
